---
title: Introduction to R
author: Data Carpentry contributors
minutes: 45
---


```{r, echo=FALSE, purl=FALSE, message = FALSE}
source("setup.R")
```

------------


------------

OK guys, we are going to move on to the second part of the lesson now, which is where we are going to be running actual commands in R. I'm going to start a new R script for this one by clicking at the upper left here. I'm going to paste in the learning objectives for this part of the session.

```{r, eval=FALSE, purl=FALSE}
#> ### Learning Objectives
#>
#> * Define the following terms as they relate to R: object, assign, call,
#>   function, arguments, options.
#> * Assign values to objects in R.
#> * Learn how to _name_ objects
#> * Use comments to inform script.
#> * Solve simple arithmetic operations in R.
#> * Call functions and use arguments to change their default options.
#> * Inspect the content of vectors and manipulate their content.
#> * Subset and extract values from vectors.
#> * Analyze vectors with missing data.
```


Alright, let's get going! Please try to follow along in Rstudio, and if you get stuck, put up a stickie so one of the awesome helpers can give you a hand. I will be copying all this code into our etherpad, so you don't need to worry about copying the explanations that start with a # symbol, just try to run the actual commands we shoot down to the console.

To start off, You can get output from R simply by typing math:

```{r, purl=FALSE}
# R can do maths for you
# Ctrl+Enter sends your commands from here, the script window, to the console below
3 + 5
12 / 7
```

And right there we've done one of our learning objectives!
```{r, purl=FALSE}
#Solve simple arithmetic operations in R.
```

Stickies if you couldn't get that to work

However, most of the time we are doing more than simple arithmetic and might need to save these values. To do this we need to assign _values_ to _objects_

```{r, purl=FALSE}
# To assigh a value to an object you use this guy "<-", the assignment operator
# It assign values on the right to objects on the left
# object <- value
weight_kg <- 55
# Remember Ctrl+Enter shoots stuff down to the console below
```

You'll might see = used in other people's code instead of the assignment operator becauuse ths used to be a more common way to do things, but our funny little arrow guy is now the preferred way to do things. 

```{r, purl=FALSE}
# So that gets rid of another one of our learning objectives
# Assign values to objects in R.
```

Stickies if you are stuck

What are some good object names for us to use?

You can pretty much call them anything that doesn't start with a number
```{r, eval=FALSE, purl=FALSE}
# object names cannot start with a number 
2pac <- "not valid object name"
# Note, error messages are good! They let us know we need to fix something!
pac2 <- "this is ok though"
# ctrl+Enter to shoot things down to the console
```

However, although can call them anything, names which are meaningful and not too long are a good idea, because this will make it easier for you to follow your code when you come back to it (and for other people to understand it). Also be aware that R is case-sensitive
```{r, eval=FALSE, purl=FALSE}
# watch out for case sensitive
weight_kg <- 55
Weight_KG <- 60
# Can run our object names to see what values they contain
weight_kg
Weight_KG
```

There are a few other things to avoid naming your objects such as if, else, and  for because these are fundamental functions in R, so it won't let you defined them as something else. It's also a good idea to avoid other function names
```{r, eval=FALSE, purl=FALSE}
# This is not a good idea
mean <- "something that is not a mean function"
mean
# How do you find out if something is already a function?
mean()
thisisdefinitelynotafunction()
```

It's also a good idea to avoid dots in names because these can have special meanings in R and other languages.

```{r, eval=FALSE, purl=FALSE}
# This is also not a good idea
putting.dots.in.names <- 304
putting.dots.in.names
```

It is also recommended to use nouns for object names, and verbs for function names as part of having a clear coding style. You can read more about consistency in your coding style at the links embedded in the lesson link, which is on etherpad. 

```{r, eval=FALSE, purl=FALSE}
# And with that, we've covered another learning objective!
# Learn how to _name_ objects
```

If anything is tripping you up so far please chuck up a stickie so a helper can help you.

Just as a quick side bar, althoguh sometimes variable and objects can mean different things, in this lesson we'll use the two terms interchangeably, so if I flip-flop between the two words, that's why.

OK, back to assigning values. When you assigning a value to an object, R does not print anything by default but you can force R to print the value by using parentheses or by typing the object name:

```{r, eval=FALSE, purl=FALSE}
weight_kg <- 55    # doesn't print anything
(weight_kg <- 55)  # but putting parenthesis around the call prints the value of `weight_kg`
weight_kg          # and so does typing the name of the object
```

Now that R has `weight_kg` in memory, we can do maths on it for example converting the weight in kg to weight in pounds

```{r, eval=FALSE, purl=FALSE}
# Converting weight in kg to weight in pounds
2.2 * weight_kg
```

We can also assign a new value to an object if we want 

```{r, eval=FALSE, purl=FALSE}
# Whoops, weight should actually be 57.5 not 55, let's change it!
weight_kg <- 57.5
2.2 * weight_kg
```

Changing the value assigned to an object won't change the values of objects previously calculated with it`:

```{r, eval=FALSE, purl=FALSE}
# Saving weight in pounds as a new variable, weight_lb
weight_lb <- 2.2 * weight_kg
```

```{r, eval=FALSE, purl=FALSE}
# Now changing weight_kg to 100
weight_kg <- 100
# But this doesn't change the value in weight_lb, because it was calculated using
# the old version of weight_kg
weight_lb
2.2 * weight_kg
```


### Comments
```{r, eval=FALSE, purl=FALSE}
# I've been leaving messages starting with "#" all over my scripts
# these are comments! 
# Anything to the right of the # is ignored
200*2
#300*2
```

Comments are handy for explaining what your script is doing. R studio also makes it really handy to easily comment or uncomment things
```{r, eval=FALSE, purl=FALSE}
# I can eeasily comment or uncomment all the following lines by Ctrl+Shift+c
500 * 5
500 * 4
weight_kg
weight_lb
```

```{r, eval=FALSE, purl=FALSE}
# Oh and before I forget, tab autocompletes words for you so you don't 
# make as many typos
# SHOW EXAMPLE
```

```{r, eval=FALSE, purl=FALSE}
# That's another learning objective done
#Use comments to inform script.
```

Alright, our first challenge! Going go give you guys two minutes to work through the following once I get it written down. If you get stuck, don't forget your red stickie!

```{r, eval=FALSE, purl=TRUE}
### Challenge
##
## What are the values after each statement in the following?
##
## mass <- 47.5            # mass?
## age  <- 122             # age?
## mass <- mass * 2.0      # mass?
## age  <- age - 20        # age?
## mass_index <- mass/age  # mass_index?
```

Step through the challenge

### Functions and their arguments
The next thing we are going to cover are functions, which allow us to do more complicated bits and pieces. Some functions come with R, and we can get more by importing packages, which we'll cover a bit later. 

We give the function inputs, which are known as arguments, and then it returns somehing, often a value.
```{r, eval=FALSE, purl=FALSE}
# One of the functions in R is sqrt
# sqrt takes one argument, the number we want to get the sqrt of
# the value it gives back is the sqrt of that number
sqrt(4)
# 4 is the argument, and 2 is the value returned

# If we want to save this value we need to assign it to an object
sqrt_four <- sqrt(4)
# Ctrl+Enter to shoot stuff down to the console
```

Running or Executing a function like this is also called calling the function

A function doesn't have return a number like the sqrt function does. The return value also doesn't need to be a single item:  it can be a set of things, or even a dataset. We'll see that when we read data files into R.

Arguments can also be anything, not only numbers or filenames, but also other objects. Many functions also take multiple arguments. Sometimes these functions can fill in these arguments with default values - these types of arguments are known as options. We'll talk about how to figure out what arguments functions need in a sec, using a  function that can take multiple arguments: `round()`.

```{r, results='show', purl=FALSE}
# But in the meantime we've achieved another learning objective!
# Define the following terms as they relate to R: object, assign, call,
# function, arguments, options.
```

Stickies if you are having trouble following

OK, back to round...
```{r, results='show', purl=FALSE}
round(3.14159)
```

Here, we've called `round()` with just one argument, `3.14159`, and it has returned the value `3`.  That's because the default is to round to the nearest whole number. 

```{r, results='show', purl=FALSE}
# To get more information about the arguments we can use the args function
args(round)

# Or look at the help for round using
?round
```

```{r, results='show', purl=FALSE}
# We see that if we want a different number of digits, we can
round(3.14159, digits = 2)
```

```{r, results='show', purl=FALSE}
# If you provide the arguments in the exact same order as they are defined you
# don't have to name them:
round(3.14159, 2)
```

However, It is good practice to put the non-optional arguments (like the number you're
rounding) first in your function call, and to specify the names of all optional
arguments.  If you don't, someone reading your code might have to look up the
definition of a function with unfamiliar arguments to understand what you're
doing.

```{r, results='show', purl=FALSE}
# If you name the arguments, you can switch their order:
round(digits = 2, x = 3.14159)
```

```{r, results='show', purl=FALSE}
# And that's another learning objective - 
# Call functions and use arguments to change their default options.
```

Stickies if you are stuck

## Vectors and data types

Alright now we are goign to move on and talk about data types in R.

A vector is the most common and basic data type in R, : it is composed by a series of values, which can be either numbers or characters. 

```{r, purl=FALSE}
# We can assign a series of values to a vector using the c() function
weight_g <- c(50, 60, 65, 82)
weight_g
```

```{r, eval=FALSE, purl=FALSE}
# A vector can also contain characters:
animals <- c("mouse", "rat", "dog")
animals

# The quotes around "mouse", "rat", etc. are essential
# otherwise R will assume there are objects called `mouse`, `rat` and `dog
animals <- c(mouse, rat, dog)
# Again, error messages == good! They tell us what we are doing wrong!
```

There a bunch of different ways to learn about what's going on with a given vector

```{r, eval=FALSE, purl=FALSE}
#length() tells you how many elements are in a particular vector:
length(weight_g)
length(animals)
```

```{r, eval=FALSE, purl=FALSE}
# All the elements in a vector have to have the same type of data
frankenvector <- c("rat",3,4)
# If you try and mix diferent types of data, they will all be coerced to one type

# You can use class() to look at the type of data in a vector
class(weight_g)
class(animals)
class(frankenvector)
# Reember you can use tab to autocomplete names
```

```{r, eval=FALSE, purl=FALSE}
# The function `str()` provides an overview of the structure of an object
# especially good for large and complex objects
str(weight_g)
str(animals)
str(frankenvector)
```

What are some other things you might want to get up to? You might want to add more elements to your vector. 
```{r, eval=FALSE, purl=FALSE}
# You can use the `c()` function to add other elements to your vector:
#Take the original vector `weight_g`, add the value `90` to the end of it, and save the # result back into`weight_g`.
weight_g <- c(weight_g, 90) # add to the end of the vector
# Then we add the value `30` to the beginning, again saving the result back into 
# `weight_g`.
weight_g <- c(30, weight_g) # add to the beginning of the vector
weight_g
```

We can do this over and over again to grow a vector, or assemble a dataset, perhaps as part of a function we write ourselves to collect results.

Alright so some more about vectors. the kinds of vectors we've been talking about are more precisely known as astomic vectors - vectors that have just a single data type.
-- We've covered  2 of the 6 main **atomic vector** types`"character"` and `"numeric"` (which is also known as `"double"`). 

The other 4 **atomic vector** types are:

* `"logical"` for `TRUE` and `FALSE` (the boolean data type)
* `"integer"` for integer numbers 

and two other types we won't really go into here:
* `"complex"` to represent complex numbers with real and imaginary parts (e.g.,
  `1 + 4i`) and that's all we're going to say about them
* `"raw"` for bitstreams that we won't discuss further

```{r, eval=FALSE, purl=FALSE}
# You can check the type of your vector using the `typeof()` function 
# with  your vector as the argument.
typeof(animals)
typeof(weight_g)
typeof(frankenvector)
```

Aside from vectors, other important data structures are lists (`list`), matrices (`matrix`), data frames (`data.frame`), factors (`factor`) and arrays (`array`). These will come up later as we continue to work with R.

But for now, that's another learning objective down:
```{r, eval=FALSE, purl=FALSE}
#Inspect the content of vectors and manipulate their content.
```

OK guys, we've got another two minute challenge to go here!
> ### Challenge
>
> * What will happen in each of these examples? (hint: use `class()`
>   to check the data type of your objects):
>
>     ```r
>     num_char <- c(1, 2, 3, "a")
>     num_logical <- c(1, 2, 3, TRUE)
>     char_logical <- c("a", "b", "c", TRUE)
>     tricky <- c(1, 2, 3, "4")
>     ```
>
> * Why do you think it happens?
>
> ```{text_answer, echo=FALSE, purl=FALSE}
> Vectors can be of only one data type. R tries to
> convert (coerce) the content of this vector to find a "common
> denominator" that doesn't lose any information.

## Subsetting vectors

The next thing we want to learn is how to extract bits and pieces from our vector

```{r, eval=FALSE, purl=FALSE}
#If we want to extract one or several values from a vector, we provide one
#or several indices in square brackets.
animals <- c("mouse", "rat", "dog", "cat")
animals[2]
animals[c(3, 2)]
```

```{r, eval=FALSE, purl=FALSE}
# We can also repeat the indices to create an object with more elements than the
#original one:

more_animals <- animals[c(1, 2, 3, 2, 1, 4)]
more_animals
```

R indices start at 1 because that's what human beings typically do. Languages in the C family (including C++, Java, Perl, and Python) count from 0 because that's simpler for computers to do. If you are switching back and forward between python and R, that's something to try and remember.

### Conditional subsetting
Now we are going to cover how to pull out elements from a vector that satisfy a particular condition that we are interested in.

```{r, eval=FALSE, purl=FALSE}
# `TRUE` will select the element with the same index, while `FALSE` will not:
weight_g <- c(21, 34, 39, 54, 55)
weight_g[c(TRUE, FALSE, TRUE, TRUE, FALSE)]
```

Instead of typing by hand like this example, normally our true/false vectors are the output of other functions or logical tests. 

```{r, eval=FALSE, purl=FALSE}
# If you wanted to select only the values above 50:
weight_g > 50    # will return logicals with TRUE for the indices that meet the condition
## so we can use this to select only the values above 50
weight_g[weight_g > 50]
```

```{r, eval=FALSE, purl=FALSE}
# You can combine multiple tests using `&` (both conditions are true, AND) or
#`|` (at least one of the conditions is true, OR):
weight_g[weight_g < 30 | weight_g > 50]
weight_g[weight_g >= 30 & weight_g == 21]
#`<` stands for "less than"
#`>` for "greater than"
#'>=` for "greater than or equal to", 
#`==` for "equal to" because`=` sign performs variable assignment (similar to `<-`).
```

```{r, eval=FALSE, purl=FALSE}
#When comparing vectors to vectors to see if any of the elements of the query vector 
#are found in the subhject vector, %in% is handy

animals <- c("mouse", "rat", "dog", "cat")
animals[animals == "cat" | animals == "rat"] # returns both rat and cat
animals %in% c("rat", "cat", "dog", "duck", "goat")
animals[animals %in% c("rat", "cat", "dog", "duck", "goat")]
```

```{r, eval=FALSE, purl=FALSE}
# Another learning objective
#Subset and extract values from vectors.
```


Quick challenge - two minutes, stickies if you are stuck
> ### Challenge (optional){.challenge}
>
> * Can you figure out why `"four" > "five"` returns `TRUE`?
> 
> ```{text_answer, echo=FALSE, purl=FALSE}
> When using ">" or "<" on strings, R compares their alphabetical order. 
> Here "four" comes after "five", and therefore is "greater than" it.
> ```


## Missing data

As R was designed to analyze datasets, it is good at handling missing data, which are represented as `NA`. By default,  most functions will return `NA` if the data you are working with include missing values to make it harder to overlook the cases where you are dealing with missing data.

```{r, eval=FALSE, purl=FALSE}
heights <- c(2, 4, 4, NA, 6)
mean(heights)
max(heights)
#You can add the argument `na.rm=TRUE` to calculate the result while ignoring
#the missing values.
mean(heights, na.rm = TRUE)
max(heights, na.rm = TRUE)
```

If your data include missing values, you may want to become familiar with the
functions `is.na()`, `na.omit()`, and `complete.cases()`. See below for
examples.

```{r, eval=FALSE, purl=FALSE}
## Extract those elements which are not missing values.
heights[!is.na(heights)]

## Returns the object with incomplete cases removed. The returned object is an atomic vector of type `"numeric"` (or `"double"`).
na.omit(heights)

## Extract those elements which are complete cases. The returned object is an atomic vector of type `"numeric"` (or `"double"`).
heights[complete.cases(heights)]
```

OK, folks that's the last learning objective covered!
```{r, eval=FALSE, purl=FALSE}
#Analyze vectors with missing data.
```

Recall that you can use the `typeof()` function to find the type of your atomic vector.

> ### Challenge
>
> 1. Using this vector of heights in inches, create a new vector with the NAs removed.
>
>     ```r
>     heights <- c(63, 69, 60, 65, NA, 68, 61, 70, 61, 59, 64, 69, 63, 63, NA, 72, 65, 64, 70, 63, 65)
>     ```
> 2. Use the function `median()` to calculate the median of the `heights` vector.
>
> 3. Use R to figure out how many people in the set are taller than 67 inches.
>
> ```{r, answer=TRUE}
> heights <- c(63, 69, 60, 65, NA, 68, 61, 70, 61, 59, 64, 69, 63, 63, NA, 72, 65, 64, 70, 63, 65)
> 
> # 1.
> heights_no_na <- heights[!is.na(heights)] 
> # or
> heights_no_na <- na.omit(heights)
>
> # 2.
> median(heights, na.rm = TRUE)
>
> # 3.
> heights_above_67 <- heights_no_na[heights_no_na > 67]
> length(heights_above_67)
> ```

Next up, manipulating proper data sets!

```{r, child="_page_built_on.Rmd"}
```
