---
title: Data visualization with ggplot2
author: Data Carpentry contributors
minutes: 60
---


```{r setup, echo=FALSE, purl=FALSE}
source("setup.R")
surveys_complete <- read_csv(file = "data_output/surveys_complete.csv", col_types = cols())
```

```{r, echo=FALSE, purl=TRUE}
### Data Visualization with ggplot2
```

Kia ora folks!

-- If you were in the last R workshop on Monday you learned how to filter and select

-- in this session we are going to learn to plot that data

-- Start up R studio, red sticky if any issues

-- Link to class etherpad is up on page

-- In it is the link to today's lesson if you want to revisit it

-- Feel free to leave feedback in it: I get better at teaching based on feedback

-- I'll also put all the code I write today in the etherpad, starting with the learning objectives.

# Set up for data directory
If you weren't here on Monday, inside your working directory, you'll need a folder called data_output. We'll download the "here's one I prepared earlier" surveys_complete.csv file from the ResBaz webpage.s

------------

> ### Specific learning Objectives
>
> * Producing a bunch of different kinds of plots and modifying the aesthetics like labels and colour
> * We are also going to learn about  universal plot settings and faceting
> * And combining this together we are going to be able to build complex and customized plots

--------------

```{r load-package, eval=FALSE,message=FALSE, purl=FALSE}
# Load tidyverse
library(tidyverse)
```

**`ggplot2`** is included in the **`tidyverse`** package.

```{r load-data, eval=FALSE, purl=FALSE}
# Loading in data we made last time
surveys_complete <- read_csv("data_output/surveys_complete.csv")
```

STICKIES

## Plotting with **`ggplot2`**

-- ggplot graphics are built step by step

-- adding layers in allows for extensive flexibility and customization of plots.

``` {r, eval=FALSE, purl=FALSE}
# Basic ggplot template
ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) +  <GEOM_FUNCTION>()
```

```{r, eval=FALSE, purl=FALSE}
# Binding plot to specific data using data argument
ggplot(data = surveys_complete)
```

Not much to look at

```{r, eval=FALSE, purl=FALSE}
# Defining variables by mapping them using
# the aes function e.g. x/y, colour, size etc
ggplot(data = surveys_complete, mapping = aes(x = weight, y = hindfoot_length))

# Stuff defined inside ggplot() parenthenses
# applies to the whole plot
```

At least now we have axes

Now adding actual data points known as 'geoms' e.g. points, lines, bars

```{r first-ggplot, eval=FALSE, purl=FALSE}
# To add a geom to the plot use the `+` operator
ggplot(data = surveys_complete, mapping = aes(x = weight, y = hindfoot_length)) +
  geom_point()
```


As well as 
```{r, eval=FALSE, purl=FALSE}
# for scatter plots, dot plots, etc.
geom_point()

# boxplots
geom_boxplot()

# line graphs (e.g. time series)
geom_line()  
```

The `+` in the **`ggplot2`** package is  useful because it allows you
to modify existing `ggplot` objects. 

```{r, eval=FALSE, purl=FALSE}
# Assign plot to a variable
surveys_plot <- ggplot(data = surveys_complete, mapping = aes(x = weight, y = hindfoot_length))

# Now you can add points to the plot
surveys_plot + 
    geom_point()
```

**Notes**

I mentioned before that anything you put in the `ggplot()` function can be seen by any geom layers that you add (i.e., these are universal plot settings 
- BTW 1/5 of our learning objectives down: We know how to set universal plot settings.

- You can also specify mappings for a given geom independently

- BUT when we are adding geom layers there is an important rule about where the + goes

```{r, eval=FALSE, purl=FALSE}
# This is the correct syntax for adding layers
# (+ has to be at end of the the line)
surveys_plot +
  geom_point()

# This will not add the new layer and will return an error message
surveys_plot
  + geom_point()
```

One problem with scatterplots and a ton of data: too many dots all on top of each other. One solution? hexagonal binning (colour coding the hexagon based on number of observations)

```{r, ggplot-with-plus-position, eval=FALSE, purl=FALSE}
# To use hexagonal binning with ggplot2, first install the R package `hexbin`
install.packages("hexbin")
library(hexbin)

# Then use the `geom_hex()` function:
surveys_plot +
geom_hex()
```

STICKIES

## Building your plots iteratively

We've seen the basic process for building a ggplot iteratively, as well as a few of the different plot types available. Now we are going to look at some more options we can add in.

Building plots with **`ggplot2`** is typically an iterative process. We start by
defining the dataset we'll use, lay out the axes, and choose a geom:

```{r create-ggplot-object, eval=FALSE, purl=FALSE}
# Building a plot in ggplot2 is iterative
# We first define the dataset we'll use, lay out the axes, chose a geom
ggplot(data = surveys_complete, mapping = aes(x = weight, y = hindfoot_length)) +
    geom_point()
```

Point out the + is at the end of the line

Then, we start modifying this plot to extract more information from it. For
instance, we can add transparency (`alpha`) to avoid overplotting:

```{r adding-transparency, eval=FALSE,purl=FALSE}
# Can then start modifying plot to display more information
# e.g. using transperancy to avoid overplotting
ggplot(data = surveys_complete, mapping = aes(x = weight, y = hindfoot_length)) +
    geom_point(alpha = 0.1)
```

```{r adding-colors, eval=FALSE,purl=FALSE}
# Can make the points a specific colour instead of black (the default)
ggplot(data = surveys_complete, mapping = aes(x = weight, y = hindfoot_length)) +
    geom_point(alpha = 0.1, color = "blue")
```

Mention again "optional" function arguments

What about colouring by another variable, like species?

```{r color-by-species, eval=FALSE,purl=FALSE}
# How to color points by another variable - supply this to the aes() function
ggplot(data = surveys_complete, mapping = aes(x = weight, y = hindfoot_length)) +
    geom_point(alpha = 0.1, aes(color = species_id))

# If we remove the transparancy we can see the dots a little easier
ggplot(data = surveys_complete, mapping = aes(x = weight, y = hindfoot_length)) +
    geom_point(aes(color = species_id))

# Could also specify the color as a universal setting
ggplot(data = surveys_complete, mapping = aes(x = weight, y = hindfoot_length, color = species_id)) +
    geom_point()

```

STICKIES

## Challenge time!

Five minutes

```{r challenge-scatter, eval=FALSE, purl=FALSE}
# Create a scatter plot of weight over species_id and colour points by species_id

ggplot(data = surveys_complete, mapping = aes(x = species_id, y = weight)) +
geom_point(aes(color = plot_type))
```

STICKIES
Do you guys think that is a good way to show this data?

GO TO MAIN LESSON
# UP TO HERE GOING THROUGH

## Boxplot

We can use boxplots to visualize the distribution of weight within each species:

```{r boxplot, purl=FALSE}
ggplot(data = surveys_complete, mapping = aes(x = species_id, y = weight)) +
    geom_boxplot()
```

By adding points to boxplot, we can have a better idea of the number of
measurements and of their distribution:

```{r boxplot-with-points, purl=FALSE}
ggplot(data = surveys_complete, mapping = aes(x = species_id, y = weight)) +
    geom_boxplot(alpha = 0) +
    geom_jitter(alpha = 0.3, color = "tomato")
```

Notice how the boxplot layer is behind the jitter layer? What do you need to
change in the code to put the boxplot in front of the points such that it's not
hidden?

> ### Challenges
>
> Boxplots are useful summaries, but hide the *shape* of the distribution. For
> example, if the distribution is bimodal, we would not see it in a
> boxplot. An alternative to the boxplot is the violin plot, where the shape 
(of the density of points) is drawn.
>
> - Replace the box plot with a violin plot; see `geom_violin()`.
>
> In many types of data, it is important to consider the *scale* of the
> observations.  For example, it may be worth changing the scale of the axis to
> better distribute the observations in the space of the plot.  Changing the scale
> of the axes is done similarly to adding/modifying other components (i.e., by
> incrementally adding commands). Try making these modifications:
>
> - Represent weight on the log~10~ scale; see `scale_y_log10()`.
>
> So far, we've looked at the distribution of weight within species. Try making a
> new plot to explore the distribution of another variable within each species.
>
> - Create a boxplot for `hindfoot_length`. Overlay the boxplot layer on a jitter
>   layer to show actual measurements.
>
> - Add color to the data points on your boxplot according to the plot from which
>   the sample was taken (`plot_id`).
>
> *Hint:* Check the class for `plot_id`. Consider changing the class of `plot_id` from integer to factor. Why does this change how R makes the graph?

```{r boxplot-challenge, eval=FALSE, purl=TRUE, echo=FALSE}
## Challenge with boxplots:
##  Start with the boxplot we created:
ggplot(data = surveys_complete, mapping = aes(x = species_id, y = weight)) +
  geom_boxplot(alpha = 0) +
  geom_jitter(alpha = 0.3, color = "tomato")

##  1. Replace the box plot with a violin plot; see `geom_violin()`.

##  2. Represent weight on the log10 scale; see `scale_y_log10()`.

##  3. Create boxplot for `hindfoot_length` overlaid on a jitter layer.

##  4. Add color to the data points on your boxplot according to the
##  plot from which the sample was taken (`plot_id`).
##  *Hint:* Check the class for `plot_id`. Consider changing the class
##  of `plot_id` from integer to factor. Why does this change how R
##  makes the graph?

```


## Plotting time series data

Let's calculate number of counts per year for each species. First we need
to group the data and count records within each group:

```{r, purl=FALSE}
yearly_counts <- surveys_complete %>%
                 count(year, species_id)
```

Time series data can be visualized as a line plot with years on the x axis and counts
on the y axis:

```{r first-time-series, purl=FALSE}
ggplot(data = yearly_counts, mapping = aes(x = year, y = n)) +
     geom_line()
```

Unfortunately, this does not work because we plotted data for all the species
together. We need to tell ggplot to draw a line for each species by modifying
the aesthetic function to include `group = species_id`:

```{r time-series-by-species, purl=FALSE}
ggplot(data = yearly_counts, mapping = aes(x = year, y = n, group = species_id)) +
    geom_line()
```

We will be able to distinguish species in the plot if we add colors (using `color` also automatically groups the data):

```{r time-series-with-colors, purl=FALSE}
ggplot(data = yearly_counts, mapping = aes(x = year, y = n, color = species_id)) +
    geom_line()
```

## Faceting

**`ggplot2`** has a special technique called *faceting* that allows the user to split one
plot into multiple plots based on a factor included in the dataset. We will use it to
make a time series plot for each species:

```{r first-facet, purl=FALSE}
ggplot(data = yearly_counts, mapping = aes(x = year, y = n)) +
    geom_line() +
    facet_wrap(~ species_id)
```

Now we would like to split the line in each plot by the sex of each individual
measured. To do that we need to make counts in the data frame grouped by `year`,
`species_id`, and `sex`:

```{r, purl=FALSE}
 yearly_sex_counts <- surveys_complete %>%
                      count(year, species_id, sex)
```

We can now make the faceted plot by splitting further by sex using `color` (within a single plot):

```{r facet-by-species-and-sex, purl=FALSE}
 ggplot(data = yearly_sex_counts, mapping = aes(x = year, y = n, color = sex)) +
     geom_line() +
     facet_wrap(~ species_id)
```

Usually plots with white background look more readable when printed.  We can set
the background to white using the function `theme_bw()`. Additionally, you can remove
the grid:

```{r facet-by-species-and-sex-white-bg, purl=FALSE}
 ggplot(data = yearly_sex_counts, mapping = aes(x = year, y = n, color = sex)) +
     geom_line() +
     facet_wrap(~ species_id) +
     theme_bw() +
     theme(panel.grid = element_blank())
```

## **`ggplot2`** themes

In addition to `theme_bw()`, which changes the plot background to white, **`ggplot2`**
comes with several other themes which can be useful to quickly change the look
of your visualization. The complete list of themes is available
at <http://docs.ggplot2.org/current/ggtheme.html>. `theme_minimal()` and
`theme_light()` are popular, and `theme_void()` can be useful as a starting
point to create a new hand-crafted theme.

The
[ggthemes](https://jrnold.github.io/ggthemes/reference/index.html) package
provides a wide variety of options (including an Excel 2003 theme).
The [**`ggplot2`** extensions website](https://www.ggplot2-exts.org) provides a list
of packages that extend the capabilities of **`ggplot2`**, including additional
themes.


> ### Challenge

> Use what you just learned to create a plot that depicts how the average weight
> of each species changes through the years.
> 
> ```{r average-weight-time-series, answer=TRUE, purl=FALSE}
> yearly_weight <- surveys_complete %>%
>                 group_by(year, species_id) %>%
>                  summarize(avg_weight = mean(weight))
> ggplot(data = yearly_weight, mapping = aes(x=year, y=avg_weight)) +
>    geom_line() +
>    facet_wrap(~ species_id) +
>    theme_bw()
>```


```{r, eval=FALSE, purl=TRUE, echo=FALSE}
### Plotting time series challenge:
##
##  Use what you just learned to create a plot that depicts how the
##  average weight of each species changes through the years.

```

The `facet_wrap` geometry extracts plots into an arbitrary number of dimensions
to allow them to cleanly fit on one page. On the other hand, the `facet_grid`
geometry allows you to explicitly specify how you want your plots to be
arranged via formula notation (`rows ~ columns`; a `.` can be used as
a placeholder that indicates only one row or column).

Let's modify the previous plot to compare how the weights of males and females
has changed through time:

```{r average-weight-time-facet-sex-rows, purl=FALSE}
# One column, facet by rows
yearly_sex_weight <- surveys_complete %>%
    group_by(year, sex, species_id) %>%
    summarize(avg_weight = mean(weight))
ggplot(data = yearly_sex_weight, 
       mapping = aes(x = year, y = avg_weight, color = species_id)) +
    geom_line() +
    facet_grid(sex ~ .)
```

```{r average-weight-time-facet-sex-columns, purl=FALSE}
# One row, facet by column
ggplot(data = yearly_sex_weight, 
       mapping = aes(x = year, y = avg_weight, color = species_id)) +
    geom_line() +
    facet_grid(. ~ sex)
```

## Customization

Take a look at the [**`ggplot2`** cheat sheet](https://www.rstudio.com/wp-content/uploads/2016/11/ggplot2-cheatsheet-2.1.pdf), and
think of ways you could improve the plot.

Now, let's change names of axes to something more informative than 'year'
and 'n' and add a title to the figure:

```{r number-species-year-with-right-labels, purl=FALSE}
ggplot(data = yearly_sex_counts, mapping = aes(x = year, y = n, color = sex)) +
    geom_line() +
    facet_wrap(~ species_id) +
    labs(title = "Observed species in time",
         x = "Year of observation",
         y = "Number of species") +
    theme_bw()
```

The axes have more informative names, but their readability can be improved by
increasing the font size:

```{r number-species-year-with-right-labels-xfont-size, purl=FALSE}
ggplot(data = yearly_sex_counts, mapping = aes(x = year, y = n, color = sex)) +
    geom_line() +
    facet_wrap(~ species_id) +
    labs(title = "Observed species in time",
        x = "Year of observation",
        y = "Number of species") +
    theme_bw() +
    theme(text=element_text(size = 16))
```

Note that it is also possible to change the fonts of your plots. If you are on
Windows, you may have to install
the [**`extrafont`** package](https://github.com/wch/extrafont), and follow the
instructions included in the README for this package.

After our manipulations, you may notice that the values on the x-axis are still not
properly readable. Let's change the orientation of the labels and adjust them
vertically and horizontally so they don't overlap. You can use a 90-degree
angle, or experiment to find the appropriate angle for diagonally oriented
labels:

```{r number-species-year-with-theme, purl=FALSE}
ggplot(data = yearly_sex_counts, mapping = aes(x = year, y = n, color = sex)) +
    geom_line() +
    facet_wrap(~ species_id) +
    labs(title = "Observed species in time",
        x = "Year of observation",
        y = "Number of species") +
    theme_bw() +
    theme(axis.text.x = element_text(colour = "grey20", size = 12, angle = 90, hjust = 0.5, vjust = 0.5),
                        axis.text.y = element_text(colour = "grey20", size = 12),
          text = element_text(size = 16))
```

If you like the changes you created better than the default theme, you can save them as
an object to be able to easily apply them to other plots you may create:


```{r number-species-year-with-right-labels-xfont-orientation, purl=FALSE}
grey_theme <- theme(axis.text.x = element_text(colour = "grey20", size = 12, angle = 90, hjust = 0.5, vjust = 0.5),
                          axis.text.y = element_text(colour = "grey20", size = 12),
                          text = element_text(size = 16))
ggplot(surveys_complete, aes(x = species_id, y = hindfoot_length)) +
    geom_boxplot() +
    grey_theme
```

> ### Challenge
> 
> With all of this information in hand, please take another five minutes to
> either improve one of the plots generated in this exercise or create a
> beautiful graph of your own. Use the RStudio [**`ggplot2`** cheat sheet](https://www.rstudio.com/wp-content/uploads/2016/11/ggplot2-cheatsheet-2.1.pdf)
> for inspiration. Here are some ideas:
>
> * See if you can change the thickness of the lines.
> * Can you find a way to change the name of the legend? What about its labels?
> * Try using a different color palette (see
>   http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/).

## Arranging and exporting plots

Faceting is a great tool for splitting one plot into multiple plots, but sometimes you may want to produce a single figure that contains multiple plots using different variables or even different data frames. The **`gridExtra`** package allows us to combine separate ggplots into a single figure using `grid.arrange()`:

```{r install-gridextra, message=FALSE, purl=TRUE, eval=FALSE}
install.packages("gridExtra")
```

```{r gridarrange-example, message=FALSE, purl=FALSE, fig.width=10}
library(gridExtra)

spp_weight_boxplot <- ggplot(data = surveys_complete, 
                             mapping = aes(x = species_id, y = weight)) +
  geom_boxplot() +
  xlab("Species") + ylab("Weight (g)") +
  scale_y_log10()

spp_count_plot <- ggplot(data = yearly_counts, 
                         mapping = aes(x = year, y = n, color = species_id)) +
  geom_line() + 
  xlab("Year") + ylab("Abundance")

grid.arrange(spp_weight_boxplot, spp_count_plot, ncol = 2, widths = c(4, 6))

```

In addition to the `ncol` and `nrow` arguments, used to make simple arrangements, there are tools for [constructing more complex layouts](https://cran.r-project.org/web/packages/gridExtra/vignettes/arrangeGrob.html). 

After creating your plot, you can save it to a file in your favorite format. The Export tab in the **Plot** pane in RStudio will save your plots at low resolution, which will not be accepted by many journals and will not scale well for posters. 

Instead, use the `ggsave()` function, which allows you easily change the dimension and resolution of your plot by adjusting the appropriate arguments (`width`, `height` and `dpi`). 

Make sure you have the `fig_output/` folder in your working directory.

```{r ggsave-example, eval=FALSE, purl=FALSE}
my_plot <- ggplot(data = yearly_sex_counts, 
                  mapping = aes(x = year, y = n, color = sex)) +
    geom_line() +
    facet_wrap(~ species_id) +
    labs(title = "Observed species in time",
        x = "Year of observation",
        y = "Number of species") +
    theme_bw() +
    theme(axis.text.x = element_text(colour = "grey20", size = 12, angle = 90, hjust = 0.5, vjust = 0.5),
                        axis.text.y = element_text(colour = "grey20", size = 12),
          text=element_text(size = 16))
ggsave("fig_output/yearly_sex_counts.png", my_plot, width = 15, height = 10)

# This also works for grid.arrange() plots
combo_plot <- grid.arrange(spp_weight_boxplot, spp_count_plot, ncol = 2, widths = c(4, 6))
ggsave("fig_output/combo_plot_abun_weight.png", combo_plot, width = 10, dpi = 300)
```

Note: The parameters `width` and `height` also determine the font size in the saved plot.


```{r final-challenge, eval=FALSE, purl=TRUE, echo=FALSE}
### Final plotting challenge:
##  With all of this information in hand, please take another five
##  minutes to either improve one of the plots generated in this
##  exercise or create a beautiful graph of your own. Use the RStudio
##  ggplot2 cheat sheet for inspiration:
##  https://www.rstudio.com/wp-content/uploads/2015/08/ggplot2-cheatsheet.pdf
```


**`ggplot2`** is a plotting package that makes it simple to create complex plots
from data in a data frame. It provides a more programmatic interface for
specifying what variables to plot, how they are displayed, and general visual
properties. Therefore, we only need minimal changes if the underlying data change
or if we decide to change from a bar plot to a scatterplot. This helps in creating
publication quality plots with minimal amounts of adjustments and tweaking.

**`ggplot2`** functions like data in the 'long' format, i.e., a column for every dimension,
and a row for every observation. Well-structured data will save you lots of time
when making figures with **`ggplot2`**


STICKIES
FEEDBACK IN ETHERPAD


```{r, child="_page_built_on.Rmd"}
```
