---
title: "Starting with data"
author: "Data Carpentry contributors"
minutes: 20
---

```{r, echo=FALSE, purl=FALSE, message = FALSE}
source("setup.R")
```

------------

Kia ora

-- last time we learned how to create objects and use comments and functions

-- we also learned about a data type called a vector

-- which you could think of as a bunch of values all in a line

-- in this session we are going to learn about data frames, which are another data type

-- kind of like R's version of an excel spreadsheet

-- Start up R studio, red sticky if any issues

-- Link to class etherpad is up on page

-- In it is the link to today's lesson if you want to revisit it

-- Feel free to leave feedback in it: I get better at teaching based on feedback

-- I'll also put all the code I write today in the etherpad so I think you'll get most out of following along and trying to do the code rather than taking notes

> ### Specific Learning Objectives
>
> * You'll be learning about data frames, how to pull a comma-delimited-file (an csv file) from your computer into R to create a data frame, and how to figure out what is in the data frame.
> * We'll also be talking a lot about something called a factor, how to convert between strings and factors, and how to handle strings and factors
> * We'll finish up with how to format dates.

------------

STICKIES

## Getting the data

```{r, eval=FALSE, purl=TRUE}
# We are going to use the R function `download.file()` to download the CSV file
# download.file("where are we downloading from?", "what do we want to call the downloaded file?")

download.file("https://ndownloader.figshare.com/files/2292169",
              "data/portal_data_joined.csv")
```

This is naming the data "portal_data_joined.csv," and adding it to a preexisting folder named "data."

Problems: if you don't have a folder called data you will get an error:

error in download.file cannot open destfile reason 'no such file or directory'

We now have the data on our computer, but no we need to get it into R

```{r, eval=TRUE,  purl=FALSE}
#we use `read.csv()` to load our object into memory as a `data.frame`. 
surveys <- read.csv("data/portal_data_joined.csv")
```

This statement doesn't produce any output because, as you might recall,
assignments don't display anything. We could put brackets about it....

```{r, eval=TRUE,  purl=FALSE}
#we use `read.csv()` to load our object into memory as a `data.frame`. 
(surveys <- read.csv("data/portal_data_joined.csv"))

# Or we check that our data has been loaded by typing its name
surveys
```

Phew boy that is a lot of output.

```{r, results='show', purl=FALSE}
# To look at just the first 6 lines of this data frame
head(surveys)
# Try also
View(surveys)
```

view is handy because you can sort things

STICKIES

## Presentation of the Survey Data

So what is actually in this data set?

We are studying the species  and weight of animals caught in plots in our study

Each row holds information for a single animal, and the columns represent:

| Column           | Description                        |
|------------------|------------------------------------|
| record\_id       | Unique id for the observation      |
| month            | month of observation               |
| day              | day of observation                 |
| year             | year of observation                |
| plot\_id         | ID of a particular plot            |
| species\_id      | 2-letter code                      |
| sex              | sex of animal ("M", "F")           |
| hindfoot\_length | length of the hindfoot in mm       |
| weight           | weight of the animal in grams      |
| genus            | genus of animal                    |
| species          | species of animal                  |
| taxon             | e.g. Rodent, Reptile, Bird, Rabbit |
| plot\_type       | type of plot                       |


```{r, eval=FALSE,  purl=FALSE}
# A few notes on data input...

# Fields delimited by commas
read.csv()

# However in some countries, "," is the decimal point and ; separates columns
read.csv2()

# There is also the following for tab separated data files
read.delim()

# Or if you have other characters delimiting your fields
read.table()
```

All of these functions are actually wrapper functions for read.table

```{r, eval=FALSE,  purl=FALSE}
# Therefore, another way we could have loaded our data was:
`surveys <- read.table(file="data/portal_data_joined.csv", sep=",", header=TRUE)`.
```

The header argument has to be set to TRUE to be able to read the headers as the default for the header option is FALSE.

STICKIES

## So what IS a data frames?

A data frame is a table where the columns are vectors that all have the same length. 

Because columns are vectors, each column must contain a single type of data (e.g., characters, integers, factors).

We can see this in our data by:

```{r, purl=FALSE}
# str() will show the type of daa for each of our column vectors
str(surveys)
```

Data frames are most commonly they are generated by read.table() or other similar function and are the _de facto_ data structure for most tabular data, and useful for statistics and plotting.

Guys - that's 2 out of our 8 learning objectives down

Being able to describe what a data frame is

Loading data in from a *csv into a data frame.

Nice work!

STICKIES

## Inspecting `data.frame` Objects

OK, now we are going to go over some ways to check the content/structure of the data

```{r, purl=FALSE}
### Looking at size ###
# dim() gives a vector with the number of rows in the first element
# and the number of columns as the second element. dim = dimensions
dim(surveys)

# number of rows
nrow(surveys)

# number of columns
ncol(surveys)

### Looking at the content ###
# Show the first 6 rows
head(surveys)
# Show the last 6 rows
tail(surveys)

### Looking at names ####
# Returns column names 
names(surveys)
# Synonym - for data.frame
colnames(surveys)
# Return row names
rownames(surveys)

### Summary ###
# Structure of object and class, length and content of each column
str(surveys)
# summary statistics for each column
summary(surveys)
```

STICKIES

### Challenge time

Five minutes

```{r, purl=FALSE}
#Based on the output of `str(surveys)`, can you answer the following questions?

# What is the class of the object `surveys`?
# How many rows and how many columns are in this object?
# How many species have been recorded during these surveys?
```

```{r, answer=TRUE, results="markup", purl=FALSE}
str(surveys)
# class: data frame
# how many rows: 34786,  how many columns: 13
# how many species: 48
```

And boom! That's our 3rd out of eight learning objectives down:

Because now we now know how to summarize the contents of a data frame

STICKIES

## Indexing and subsetting data frames

```{r, eval=FALSE,echo=FALSE, purl=TRUE}
# When extracting a 1-dimensional vector, only have to give one coordinate e.g.
# To extract the fifth element of my_vector:
my_vector[5]

```

Because a data frame has 2 dimensions we need to specify the "coordinates" we want from it. However, note that different ways of specifying these coordinates lead to results with different classes.


```{r, purl=FALSE}
# When doing coordinates: row numbers come first, followed by column numbers. 

# first element in the first column of the data frame (as a vector)
surveys[1, 1]   
# first element in the 6th column (as a vector)
surveys[1, 6]   
# first column of the data frame (as a vector)
surveys[, 1]    
# first column of the data frame (as a data.frame)
surveys[1]      
# `:` is a special function that creates numeric vectors of integers 
# e.g. `1:3` increasing, and `3:1` decreasing
# e.g. to get first three elements in the 7th column (as a vector)
surveys[1:3, 7] 
# the 3rd row of the data frame (as a data.frame)
surveys[3, ]    
# equivalent to head_surveys <- head(surveys)
head_surveys <- surveys[1:6, ] 
```

```{r, purl=FALSE}
# You can also exclude certain indices of a data frame using the "`-`" sign:
surveys[, -1]          # The whole data frame, except the first column
surveys[-c(7:34786), ] # Equivalent to head(surveys)
```

Data frames can be subset by calling indices (as shown previously), but also by calling their column names directly:

```{r, eval = FALSE, purl=FALSE}
surveys["species_id"]       # Result is a data.frame
surveys[, "species_id"]     # Result is a vector
surveys[["species_id"]]     # Result is a vector
surveys$species_id          # Result is a vector
```

In RStudio, you can use the autocompletion feature to get the full and correct names of the columns.

STICKIES

### Challenge time

Five minutes

```{r, eval = FALSE, purl=FALSE}
# 1. Create a `data.frame` (`surveys_200`) containing only the data in
# row 200 of the `surveys` dataset.

# 2. Notice how `nrow()` gave you the number of rows in a `data.frame`?
# Use that number to pull out just that last row in the data frame.
# Compare that with what you see as the last row using `tail()` to make sure it's meeting expectations.
# Pull out that last row using `nrow()` instead of the row number.
# Create a new data frame (`surveys_last`) from that last row.

# 3. Use `nrow()` to extract the row that is in the middle of the data
# frame. Store the content of this row in an object named `surveys_middle`.

# 4. Combine `nrow()` with the `-` notation above to reproduce the behavior of
# `head(surveys)`, keeping just the first through 6th rows of the surveys
# dataset.
```

```{r, answer=TRUE, purl=FALSE}
## 1.
surveys_200 <- surveys[200, ]
## 2.
# Saving `n_rows` to improve readability and reduce duplication
n_rows <- nrow(surveys)
surveys_last <- surveys[n_rows, ]
## 3.
surveys_middle <- surveys[n_rows / 2, ]
## 4.
surveys_head <- surveys[-(7:n_rows), ]
```

STICKIES

## Factors

```{r, echo=FALSE, purl=TRUE}
# What are factors?
str(surveys)
```

Columns consist of different data types, including factors

Factors represent categories of data. This allows R to store the data as integers tied to these categories.

While they LOOK like character vectors, they are actually treated as integer vectors by R which can be kind of tricky

Once created, factors can only contain a pre-defined set of values, known as
*levels*. This is kind of like our spreadsheet QC section with the drop down options for what we allowed in a cell

```{r, purl=TRUE}
# By default, R always sorts levels in alphabetical order e.g.
sex <- factor(c("male", "female", "female", "male"))
```

R will assign `1` to the level `"female"` and `2` to the level `"male"` (because
`f` comes before `m`, even though the first element in this vector is
`"male"`). 

```{r, purl=TRUE}
# Looking at order of levels
levels(sex)

# How any levels do I have?
nlevels(sex)
```

Sometimes, you want to change the order of the factors because that makes more sense e.g., "low", "medium", "high"
it improves your visualization, or it is required by a particular type of
analysis. Here, one way to reorder our levels in the `sex` vector would be:

```{r, results=TRUE, purl=FALSE}
# How to change the order of the factors
sex # current order
sex <- factor(sex, levels = c("male", "female"))
sex # after re-ordering
```

Even though in R's memory, these factors are represented by integers (1 and 2) they are more informative because we know 1 is female and 2 is male. 

STICKIES

### Converting factors

Converting factors

```{r, purl=FALSE}
# If you need to convert a factor to a character vector, you use `as.character(x)`.
as.character(sex)
```

Converting factors where the levels appear as numbers (such as concentration
levels, or years) to a numeric vector is a little trickier. 

```{r, purl=TRUE}
year_fct <- factor(c(1990, 1983, 1977, 1998, 1990))
year_fct
# The `as.numeric()` function returns the index values of the factor, not its levels, 
# so it will result in an entirely new (and unwanted in this case) set of numbers. 
as.numeric(year_fct)               # Wrong! And there is no warning...

as.numeric(as.character(year_fct)) # Works...
as.numeric(levels(year_fct))[year_fct]    # The recommended way.

# In the `levels()` approach, three important steps occur:
#1. We obtain all the factor levels using `levels(year_fct)`
#2. We convert these levels to numeric values using `as.numeric(levels(year_fct))`
#3. We then access these numeric values using the underlying integers of the
# vector `year_fct` inside the square brackets
```

STICKIES

### Renaming factors

```{r, purl=TRUE}
# When your data is stored as a factor, you can use the `plot()` function to get a
# quick glance at it

## bar plot of the number of females and males captured during the experiment:
plot(surveys$sex)
```

In addition to males and females, there are about 1700 individuals for which the sex information hasn't been recorded and we don't know if this information is missing or undetermined.


```{r, results=TRUE, purl=FALSE}
# Pulling out the data on sex and work with that data, so we're not modifying the 
# original data frame:

sex <- surveys$sex
head(sex)
levels(sex)
levels(sex)[1] <- "undetermined"
levels(sex)
head(sex)
plot(sex)
# Now we have labels for our "undetermined" columns
```

STICKIES

Folks! That's another three objectives we've knocked off! We now know how to:
- Describe what a factor is.
- Convert between strings and factors.
- Reorder and rename factors.

### Challenge

Five minutes

```{r, results=TRUE, purl=FALSE}
# Rename "F" and "M" to "female" and "male" respectively.
# Now that we have renamed the factor level to "undetermined", can you recreate the
# barplot such that "undetermined" is last (after "male")?
```

```{r, answer=TRUE, purl=FALSE}
levels(sex)[2:3] <- c("female", "male")
sex <- factor(sex, levels = c("female", "male", "undetermined"))
plot(sex)
``` 

STICKIES

### Using `stringsAsFactors=FALSE`

By default, when making a data frame, columns that contain characters (i.e. text) are coerced (= converted) into factors. 

You may (I would say often) want to keep these columns as `character`. To do so, in `read.csv()` and `read.table()` there is an optionalargument called `stringsAsFactors` which can be set to `FALSE`.

```{r, eval=FALSE, purl=TRUE}
## Compare the difference between our data read as `factor` vs `character`.
surveys <- read.csv("data/portal_data_joined.csv", stringsAsFactors = TRUE)
str(surveys)
surveys <- read.csv("data/portal_data_joined.csv", stringsAsFactors = FALSE)
str(surveys)

# If we decide we DO want factors, we can convert the column into a factor
surveys$plot_type <- factor(surveys$plot_type)
```

STICKIES

And boom, second to last learning objective for this lesson covered! We now know how character strings are handled in a data frame.

### Challenge

FIVE MINUTES

```{r, eval=FALSE, purl=TRUE}
# 1. We have seen how data frames are created when using `read.csv()`, but
# they can also be created by hand with the `data.frame()` function.  There are
# a few mistakes in this hand-crafted `data.frame`. Can you spot and fix them?
# Don't hesitate to experiment!

animal_data <- data.frame(animal = c(dog, cat, sea cucumber, sea urchin),feel = c("furry", "squishy", "spiny"),weight = c(45, 8 1.1, 0.8))

#2. Can you predict the class for each of the columns in the following example?

country_climate <- data.frame(country = c("Canada", "Panama", "South Africa", "Australia"),climate = c("cold", "hot", "temperate", "hot/temperate"),temperature = c(10, 30, 18, "15"),northern_hemisphere = c(TRUE, TRUE, FALSE, "FALSE"),has_kangaroo = c(FALSE, FALSE, FALSE, 1))

#Check your guesses using 
str(country_climate)

#Are they what you expected?  Why? Why not?
# What would have been different if we had added `stringsAsFactors = FALSE` when creating the data frame?
# What would you need to change to ensure that each column had the accurate data type?
```

ANSWERS
```{r, eval=FALSE, purl=TRUE}
# 1
# missing quotations around the names of the animals
# missing one entry in the "feel" column (probably for one of the furry animals)
# missing one comma in the weight column

# 2
# `country`, `climate`, `temperature`, and `northern_hemisphere` are factors; `has_kangaroo` is numeric using `stringsAsFactors = FALSE` would have made them character instead of factors
# removing the quotes in temperature and northern_hemisphere and replacing 1 by TRUE in the `has_kangaroo` column would give what was probably intended
```

The automatic conversion of data type is sometimes a blessing, sometimes an
annoyance. Be aware that it exists, learn the rules, and double check that data you import in R are of the correct type within your data frame. If not, use it to your advantage to detect mistakes that might have been introduced during data entry (a letter in a column that should only contain numbers for instance).

Learn more in this [RStudio tutorial](https://support.rstudio.com/hc/en-us/articles/218611977-Importing-Data-with-RStudio)

STICKIES

## Formatting Dates

Anyone from America here?

Dates are one of the biggest pains to deal with. Not only are there different orders, days-months-years vs month-day-year, or even year-month-day, but dates are a mix of numbers and characters.


```{r, eval=FALSE, purl=FALSE}
# Using `str()` we see our data frame has a separate column for day, month, and year, and that each contains integer values.

str(surveys)
```
We are going to use the `ymd()` function from the package **`lubridate`** (which belongs to the **`tidyverse`**)

When you load  the **`tidyverse`** (`library(tidyverse)`), most packages get loaded but **`lubridate`** is an optional extra 

```{r load-package, message=FALSE, purl=FALSE}
# library() checks out the package we want
library(lubridate)
```

`ymd()` is a function from the lubridate package that takes a vector representing year, month, and day, and converts it to a `Date` vector. `Date` is a class of data recognized by R as being a date.

We could format our date any which way  but, as a best practice, is a character vector formatted as "YYYY-MM-DD".

```{r, purl=FALSE}
# Creating a data vector
my_date <- ymd("2015-01-01")
str(my_date)
```

```{r, purl=FALSE}
#If we paste the year, month, and day separately - we get the same result
# sep indicates the character to use to separate each component
my_date <- ymd(paste("2015", "1", "1", sep = "-")) 
str(my_date)
```

Now we apply this function to the surveys dataset. 

```{r, purl=FALSE}
#Create a character vector from the `year`, `month`, and `day` columns of`surveys` using `paste()`:
paste(surveys$year, surveys$month, surveys$day, sep = "-")
```


```{r, purl=FALSE}
#Use this character vector can be used as the argument for `ymd()`:
ymd(paste(surveys$year, surveys$month, surveys$day, sep = "-"))
```

```{r, purl=FALSE}
#We can add `Date` vector to `surveys` as a new column called `date`:
surveys$date <- ymd(paste(surveys$year, surveys$month, surveys$day, sep = "-"))
str(surveys) # notice the new column, with 'date' as the class
```

Let's make sure everything worked correctly. 

```{r, results=TRUE, purl=FALSE}
# One way to inspect the new column is to use `summary()`:
summary(surveys$date)
```

Something went wrong: some dates have missing values. Let's investigate where they are coming from.

-- use formulas we've already seen for dealing with NAs

-- to find the rows that are no bueno


```{r, results=TRUE, purl=FALSE}
# Finding records that have `NA` in our new column `date`.
is_missing_date <- is.na(surveys$date)

# Telling R what the date columns are in our dataset
date_columns <- c("year", "month", "day")

#extract the columns "year, "month", "day" from the spreadsheet for the records with missing data

missing_dates <- surveys[is_missing_date,  date_columns]

# We will also use `head()` so we don't clutter the output:
head(missing_dates)
```

FIVE MINUTES

```{r, results=TRUE, purl=FALSE}
# Why did these dates fail to parse? 

#If you had to use these data for your analyses, how would you deal with this situation?
```

And that's that lesson done folks! You've covered the last learnign objective for this lesson which is how to format dates

STICKIES/MINUTE CARDS

```{r, child="_page_built_on.Rmd"}
```
